% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\author{}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}

\hypertarget{intro}{%
\section{Intro}\label{intro}}

\hypertarget{recommended-literature}{%
\subsubsection{Recommended Literature}\label{recommended-literature}}

{[}\protect\hyperlink{recommended-literature}{Recommended Literature}{]}

\begin{itemize}
\tightlist
\item
  historical context
\item
  where this theorem comes from
\item
  what it does
\item
  why we care
\item
  how we're gonna go about it
\end{itemize}

\hypertarget{initial-example}{%
\subsection{initial example}\label{initial-example}}

{[}{[}motivating example{]}{]} action of SL2 on both the upper half
plane and \(\hat{R}\). put pictures.

\hypertarget{setup}{%
\section{setup}\label{setup}}

what we have in our environment: Hilbert Spaces, which representations,
etc. will need to put some parts of measure theory in here, but not sure
what is didactically good. at some point it won't matter, just that it's
in there.

copy from Kerr and Li, mostly {[}{[}setup from Kerr Li{]}{]}

\hypertarget{heavy-handed-section-about-direct-integral}{%
\subsection{heavy handed section about direct
integral}\label{heavy-handed-section-about-direct-integral}}

idk wtf I'm gonna do for that but I'll figure it out. (oh me, of high
hopes.) {[}{[}Unitary Representations and Direct Integral{]}{]}

\hypertarget{proof-of-theorem}{%
\section{Proof of Theorem}\label{proof-of-theorem}}

Put a prelude in here, reminding people what we're doing here as we've
just distracted them with the setup garbage

\hypertarget{start-with-proof-using-sl2}{%
\subsection{start with proof using
SL2}\label{start-with-proof-using-sl2}}

{[}{[}the SL(2, R) part of the proof{]}{]}

\hypertarget{generalize-to-sln}{%
\subsection{generalize to SLn}\label{generalize-to-sln}}

{[}{[}The SL(n, R) part of the proof{]}{]} {[}{[}The proof for a general
G{]}{]}

\hypertarget{outlook}{%
\section{Outlook}\label{outlook}}

return to initial example.

\hypertarget{appendices}{%
\section{Appendices}\label{appendices}}

put any garbage I assume known in here? -\textgreater{} prob not. needs
to precede the proof

\hypertarget{functional-analysis}{%
\subsection{Functional analysis}\label{functional-analysis}}

\hypertarget{representation-theory}{%
\subsection{Representation Theory}\label{representation-theory}}

\hypertarget{direct-integral}{%
\subsection{Direct Integral}\label{direct-integral}}

\end{document}
